Visualizing Independence Using Extended Association Plots

### 2005-08-04 trunk/vcd/paper/JCGS.Rnw
section 3 finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@430 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\textbf{Key Words:} Association plots; Conditional inference; 
  Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
\end{abstract}

@@

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis (\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating from the permutation distribution of test statistics of type
$\lambda(\{r_{ij}\})$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistion from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
by drawing a sample of size $5000$ from the permutation distribution under independence.

@@

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
of $2$ and $4$ respectively. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette). By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

The significance $alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
Of course, users could choose any other set of significance levels they feel comfortable
with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
etc. Another option could be to use a continuous shading where the $p$ value corresponding
to a cell controls the interpolation between the neutral and the full color. However, our
experience is that this typically results in too much color in a plot which in turn tends
to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
to a shading with fewer colors that are easier to interpret.

This extended shading is already very flexible and combines visualization and inference,
but it should only be applied when using the maximum statistic is appropriate. For 
the sum-of-squares statistic $X^2$ (and all other functionals $\lambda(\cdot)$), this
shading cannot be used.

use pre-determined (or manually adjusted) cut offs, but visualize the significance
by the amount of color in the plot, i.e., use only little color if the corresponding
test result is non-significant.

For HCL, the amount of color can easily be controlled by varying the maximum
chroma value. So far $100$, now $20$.

For HSV, a similar effect can be achieved by varying the value dimension which
was not used up to now. Setting value to $v = 0.5$ a darker palette of colors can be
obtained.

We have seen that this approach performs very well when the test statistic
which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
it is difficult to extend this approach to general tests for independence,
in particular in the HCL space, where the ranges of chroma and luminance are
not independent. In contrast the HSV space has the advantage that all
three dimensions hue, saturation and value have the range $[0,1]$ and can
be varied independently. Therefore, a different approach than the one
described above could be to use the hue for the sign of the residuals as
before, the saturation for the absolute size of the residuals, and the
value as an indicator for the significance of some test statistic (only
using the full value when the overall test for independence rejects
the null hypothesis). The resulting mosaic plots under this paradigm
are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
clearly be seen that despite some large residuals there is no
evidence against independence for the Bundesliga data, but that the
null hypothesis has to be rejected for the Hair-Eye-Color data.

@@

@@

<<graphics>>
## load packages and data
library(vcd)
data(Arthritis)
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
data(UCBAdmissions)
ucb <- aperm(UCBAdmissions)

@@

## new max shading with HCL colors
mosaic(art, gp = shading_Friendly)
set.seed(1234)
mosaic(art, gp = shading_max)

## conditional independence
set.seed(1234)
cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
## which uses the double max functional
set.seed(1234)
coindep_test(ucb, "Dept")

## alternatively use Peason functional
## via simulation of conditional distribution:
set.seed(1234)
coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)
## via asymptotic distribution:
fm <- loglm(~ Dept * (Gender + Admit), data = ucb)
fm
plot(fm, labeling_args = list(abbreviate = c(Admit = 2)))
@

>>>>

\noindent \textbf{Key Words:} Association plots; Conditional inference; 
Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
\end{abstract}

@@

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating from the permutation distribution of test statistics of type
$\lambda(\{r_{ij}\})$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from (\ref{eq:ChiSquare})
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistion from (\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $5000$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
of $2$ and $4$ respectively. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

The significance $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
Of course, users could choose any other set of significance levels they feel comfortable
with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
etc. Another option could be to use a continuous shading where the $p$ value corresponding
to a cell controls the interpolation between the neutral and the full color. However, our
experience is that this typically results in too much color in a plot which in turn tends
to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
to a shading with fewer colors that are easier to interpret.

This extended shading is already very flexible and combines visualization and inference.
However, it can only be applied when employing the maximum statistic because it is the only
aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
its critical value (independent from the values of the remaining residuals) is equivalent
to the a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_alpha) = P(\max_{i,j} |r_{ij}| \le c_alpha)$.}
Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
analysis, but it would be desirable to also have a residual-based shading that can incorporate
visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
individual cells differently but can only be realized by using different colors for the
whole table. A outlined before, colorfulness is often intuitively matched with interestingness,
therefore a rather natural idea is to use the fully colored palette only when the corresponding
test is significant and to use a less colorful palette if not. For the HCL scheme, the
amount of color can easily be controlled by varying the maximum chroma value used. For the full
colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
is now reduced to $20$ for the non-significant palette. This palette still codes the absolute
size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
for small `un-interesting' residuals, codes positive and negative residuals by different hues,
but gives less emphasis to the pattern by making the plot less colorful. A similar effect
can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
preferable.

@@

We have presented two approaches for combining visualization and significance testing
via residual-based shadings. The first one alwyas uses the fully colored palette but relies on
data-driven cut offs such that the presence of color is equivalent to significance of the
associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
codes the result of the significance test by using full colors only for significant results and
the same type of palette but with a reduced amount of color for non-significant results.

@@

<<graphics>>
## load packages and data
library(vcd)
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
ucb <- aperm(UCBAdmissions)

@@

## new max shading with HCL colors
mosaic(art, gp = shading_Friendly(lty = 1, eps = NULL))
set.seed(1071)
mosaic(art, gp = shading_max)

## conditional independence
set.seed(1071)
cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
## which uses the double max functional
set.seed(1071)
coindep_test(ucb, "Dept")

## alternatively use Peason functional
## via simulation of conditional distribution:
set.seed(1071)
coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)
## via asymptotic distribution:
fm <- loglm(~ Dept * (Gender + Admit), data = ucb)
fm
plot(fm, labeling_args = list(abbreviate = c(Admit = 2)))
@
